---
title: "Getting started with veryMAD"
description: >
  A small observation-level QC workflow for sequencing metadata.
vignette: >
  %\VignetteIndexEntry{Getting started with veryMAD}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

`veryMAD` annotates observations using robust, median absolute deviation (MAD)
thresholds. You choose the metrics, the direction of a poor value, and any
transformation. The input rows stay in the result.

```{r}
#| label: setup
#| include: false
library(veryMAD)
```

## A basic QC workflow

The example uses a small sequencing metadata table. Low library size and low
gene detection suggest poor complexity; high mitochondrial content is also a
common QC concern.

```{r}
sample_metadata <- data.frame(
  sample = paste0("sample_", 1:8),
  library_size = c(2.4e6, 2.5e6, 2.6e6, 2.7e6, 2.8e6, 2.6e6, 0.45e6, 2.5e6),
  detected_genes = c(12000, 12500, 13100, 12800, 13400, 12600, 1500, 12300),
  pct_mito = c(.04, .05, .06, .05, .07, .06, .08, .29)
)

# Describe which tail is a potential QC problem for each metric
qc_directions <- c(
  library_size = "lower",
  detected_genes = "lower",
  pct_mito = "upper"
)

# Add one flag per metric and a combined flag to the metadata
qc_annotated <- mad_qc(
  sample_metadata,
  metrics = qc_directions,
  verbose = FALSE
)
```

The result keeps all eight samples and adds logical flag columns. Inspect the
columns that matter for a quick review:

```{r}
qc_annotated[, c("sample", "library_size_mad_outlier",
                 "detected_genes_mad_outlier", "pct_mito_mad_outlier",
                 "mad_qc_outlier")]
```

`mad_qc_outlier` is `TRUE` when at least one selected metric is flagged. veryMAD
does not remove those rows, so the decision about filtering or follow-up stays
with the analysis workflow.

## Inspect thresholds with a report

Use `output = "report"` when you want thresholds and flags in a compact object
instead of adding columns to the input table.

```{r}
qc_report <- mad_qc(
  sample_metadata,
  metrics = qc_directions,
  output = "report",
  verbose = FALSE
)

qc_report$thresholds[, c("metric", "direction", "lower_raw",
                         "upper_raw", "status")]
qc_report$flags[, c("id", "mad_qc_outlier")]
```

The threshold table records the calculation status for each metric. Missing or
undersized metrics remain visible there instead of being silently turned into
a filtering decision.
