---
title: "Getting started with veryMAD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with veryMAD}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(veryMAD)
```

`veryMAD` accepts observation metadata and requires users to select numeric
metrics, directions, and transformations explicitly. This example uses a
small synthetic bulk RNA-seq cohort with 36 libraries. The values are only
illustrative, but the workflow is the same for a real metadata table.

```{r}
set.seed(101)
n_libraries <- 36L
bulk_metadata <- data.frame(
  sample_id = sprintf("bulk_%02d", seq_len(n_libraries)),
  library_size = round(rlnorm(n_libraries, log(2.5e6), 0.18)),
  mapping_rate = pmin(.995, pmax(.70, rnorm(n_libraries, .94, .015))),
  duplication_rate = pmin(.95, pmax(.10, rnorm(n_libraries, .35, .04)))
)
bulk_metadata$library_size[c(2, 11)] <- c(180000, 220000)
bulk_metadata$mapping_rate[c(5, 24)] <- c(.70, .73)
bulk_metadata$duplication_rate[31] <- .82
rownames(bulk_metadata) <- bulk_metadata$sample_id

bulk_annotated <- mad_qc(
  bulk_metadata,
  metrics = c(
    library_size = "lower",
    mapping_rate = "lower",
    duplication_rate = "upper"
  ),
  transform = c(library_size = "log1p"),
  min_n = 10,
  verbose = FALSE
)

bulk_annotated[bulk_annotated$mad_qc_outlier %in% TRUE,
  c("sample_id", "library_size_mad_outlier", "mapping_rate_mad_outlier",
    "duplication_rate_mad_outlier", "mad_qc_outlier")]
```

The annotation preserves all 36 observations and adds one logical column per
metric, plus `mad_qc_outlier`. The combined flag is `TRUE` when any selected
metric is flagged; the input rows are never filtered.

Use `output = "report"` when thresholds and observation IDs should be kept in
a compact object rather than added to the metadata table:

```{r}
bulk_report <- mad_qc(
  bulk_metadata,
  metrics = c(
    library_size = "lower",
    mapping_rate = "lower",
    duplication_rate = "upper"
  ),
  transform = c(library_size = "log1p"),
  output = "report",
  min_n = 10,
  verbose = FALSE
)

bulk_report$thresholds[, c("metric", "direction", "transform", "lower_raw",
                           "upper_raw", "status")]
head(bulk_report$flags)
```
