---
title: "Robust MAD scaling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust MAD scaling}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r}
library(veryMAD)

set.seed(303)
gene_scores <- rlnorm(120, meanlog = 4, sdlog = .8)
names(gene_scores) <- sprintf("gene_%03d", seq_along(gene_scores))
scaled_scores <- mad_scale(gene_scores)
head(round(scaled_scores, 2))
```

The vector example has 120 values, allowing the robust centre and scale to be
estimated from a more realistic feature set. For matrices, `margin = 1`
scales rows and `margin = 2` scales columns. The next example uses a 200 by 24
expression matrix and includes one missing value to demonstrate that missing
values remain missing.

```{r}
expression <- matrix(
  rlnorm(200 * 24, meanlog = 4, sdlog = .8),
  nrow = 200,
  ncol = 24,
  dimnames = list(
    sprintf("gene_%03d", 1:200),
    sprintf("sample_%02d", 1:24)
  )
)
expression[, c(4, 19)] <- expression[, c(4, 19)] * c(2.5, .5)
expression[12, 7] <- NA_real_

scaled_genes <- mad_scale(expression, margin = 1)
scaled_samples <- mad_scale(expression, margin = 2)

dim(scaled_genes)
dimnames(scaled_genes)[[1]][1:3]
round(scaled_genes[1:5, 1:5], 2)
round(scaled_samples[1:5, 1:5], 2)
is.na(scaled_genes[12, 7])
```

Dimensions and dimnames are preserved; numeric data frames are returned as
matrices. A MAD-based scale is robust to extreme values and differs from a
standard-deviation-based scaling. Zero-MAD margins can return zeros, `NA`, or
an error through `zero_mad`.
