---
title: "Interpreting MAD QC flags"
description: >
  Interpreting explicit MAD flags and edge-case statuses.
vignette: >
  %\VignetteIndexEntry{Interpreting MAD QC flags}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

MAD flags are statistical signals for review. They are not sample rejection
rules or biological diagnoses. The direction tells veryMAD which tail to
inspect:

- `lower` flags unusually small values;
- `upper` flags unusually large values;
- `both` flags either tail.

```{r}
#| label: setup
#| include: false
library(veryMAD)
```

## Read a flag in context

This small table contains common sequencing QC metrics. One sample has low
complexity, one has high mitochondrial content, and one has a missing value.

```{r}
qc_metadata <- data.frame(
  sample = paste0("sample_", 1:7),
  library_size = c(2.4e6, 2.5e6, 2.6e6, 2.7e6, 2.8e6, 0.5e6, NA_real_),
  detected_genes = c(12000, 12500, 13100, 12800, 13400, 1600, 12200),
  pct_mito = c(.04, .05, .06, .05, .07, .08, .30)
)

qc_result <- mad_qc(
  qc_metadata,
  metrics = c(
    library_size = "lower",
    detected_genes = "lower",
    pct_mito = "upper"
  ),
  verbose = FALSE
)

qc_result$flags[, c("id", "library_size", "detected_genes",
                    "pct_mito", "mad_qc_outlier")]
```

The missing library size remains missing in its flag column. The combined flag
is `NA` only when no metric is flagged and at least one metric could not be
evaluated. Review the individual flags and the source metadata before deciding
what to do with a sample.

## Threshold status and transformations

Count-like metrics can be easier to compare after an explicit transformation.
The threshold table keeps both the calculation-scale and raw-scale limits.

```{r}
transformed_result <- mad_qc(
  qc_metadata,
  metrics = c(library_size = "lower", pct_mito = "upper"),
  transform = c(library_size = "log1p"),
  verbose = FALSE
)

transformed_result$thresholds[, c(
  "metric", "direction", "transform", "lower_raw", "upper_raw", "status"
)]
```

`min_n` is a computational safeguard. If a metric has too few usable values,
its status is `insufficient_n`; if every value is missing, its status is
`all_missing`. These statuses make the limitation visible rather than turning
it into a silent pass or fail.

## Zero MAD and downstream decisions

A zero MAD means that the usable values do not vary enough to estimate a
non-zero robust spread. Choose the policy that matches the downstream task.

```{r}
constant_metric <- data.frame(
  sample = paste0("sample_", 1:6),
  detected_genes = rep(12000, 6)
)

zero_mad_result <- mad_qc(
  constant_metric,
  metrics = c(detected_genes = "lower"),
  zero_mad = "zero",
  verbose = FALSE
)

zero_mad_result$thresholds[, c("metric", "mad", "lower_raw", "status")]
```

Use `zero_mad = "na"` to leave the flags unresolved, or `zero_mad = "error"`
when a zero spread should stop the workflow. For stratified analyses, split
the data explicitly and call `mad_qc()` on each subset. veryMAD does not infer
batches, conditions, clusters, or biological causes.
