---
title: "Interpreting MAD QC flags"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting MAD QC flags}
  %\VignetteEngine{knitr::rmarkdown}
---

Choose `lower` for unusually small values, `upper` for unusually large values,
and `both` for either tail. Count-like metrics may benefit from explicit
`log1p`; bounded rates are often most interpretable on their raw scale.

The following synthetic single-cell metadata contains 180 cells and four
common QC metrics. A few cells are deliberately made unusual so that the
resulting annotation is easy to inspect.

```{r}
library(veryMAD)

set.seed(202)
n_cells <- 180L
sc_metadata <- data.frame(
  cell_id = sprintf("cell_%03d", seq_len(n_cells)),
  nCount_RNA = round(rlnorm(n_cells, log(4000), .45)),
  nFeature_RNA = round(rlnorm(n_cells, log(1200), .30)),
  percent_mito = pmin(.40, pmax(.005, rbeta(n_cells, 2, 18))),
  doublet_score = pmin(.99, pmax(.01, rnorm(n_cells, .12, .04)))
)
sc_metadata$nCount_RNA[c(1, 2)] <- c(140, 180)
sc_metadata$nFeature_RNA[c(1, 2)] <- c(85, 120)
sc_metadata$percent_mito[c(3, 4)] <- c(.62, .55)
sc_metadata$doublet_score[5] <- .98
rownames(sc_metadata) <- sc_metadata$cell_id

sc_annotated <- mad_qc(
  sc_metadata,
  metrics = c(
    nCount_RNA = "lower",
    nFeature_RNA = "lower",
    percent_mito = "upper",
    doublet_score = "upper"
  ),
  transform = c(nCount_RNA = "log1p", nFeature_RNA = "log1p"),
  min_n = 20,
  verbose = FALSE
)

sc_annotated[sc_annotated$mad_qc_outlier %in% TRUE,
  c("cell_id", "nCount_RNA_mad_outlier", "nFeature_RNA_mad_outlier",
    "percent_mito_mad_outlier", "doublet_score_mad_outlier",
    "mad_qc_outlier")]
```

This is an annotation step, not a filtering step. The full table remains
available in `sc_annotated`; a downstream workflow can decide how to review
or use the flagged cells.

```{r}
sc_report <- mad_qc(
  sc_metadata,
  metrics = c(
    nCount_RNA = "lower",
    nFeature_RNA = "lower",
    percent_mito = "upper",
    doublet_score = "upper"
  ),
  transform = c(nCount_RNA = "log1p", nFeature_RNA = "log1p"),
  output = "report",
  min_n = 20,
  verbose = FALSE
)

sc_report$thresholds[, c("metric", "direction", "transform", "median", "mad",
                         "lower_raw", "upper_raw", "status")]
```

`min_n` is a computational safeguard. A zero MAD is reported explicitly and
can be handled as `"na"`, `"zero"`, or `"error"`. These flags are statistical
heuristics, not biological diagnoses or automatic filtering decisions. The
overall `mad_qc_outlier` flag is always returned as a compact any-selected-
metric summary; it is `NA` when no metric is flagged but at least one metric
could not be evaluated.

For stratified analyses, split the metadata explicitly and call `mad_qc()` on
each subset. veryMAD does not infer batches, conditions, clusters, or causes.
