---
title: "Explicit observation-level QC with veryMAD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explicit observation-level QC with veryMAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(veryMAD)
```

## Data model

Rows are observations and columns are metadata or candidate QC metrics. Every
metric and direction is user-selected; veryMAD never selects numeric columns,
recognizes naming conventions, or filters observations. The compact report is a
list containing `flags`, `thresholds`, and `settings`.

## Bulk observations

```{r}
bulk <- data.frame(
  library_size = c(1e6, 30e6, 32e6, 31e6, 33e6),
  mapping_rate = c(.50, .92, .94, .93, .95),
  batch = c("a", "a", "a", "b", "b"),
  row.names = paste0("sample", 1:5)
)
metrics <- c(library_size = "lower", mapping_rate = "lower")
flagged <- mad_qc_bulk(bulk, metrics,
  transform = c(mapping_rate = "none"), verbose = FALSE)
report <- attr(flagged, "mad_qc")
report$thresholds
```

The same call accepts a matrix whose rows are observations. A gene-by-sample
expression matrix is not automatically a QC table.

```{r}
mad_qc_bulk(as.matrix(bulk[c("library_size", "mapping_rate")]), metrics,
  transform = c(mapping_rate = "none"), verbose = FALSE)
```

## Single-cell metadata

```{r}
cells <- data.frame(
  nCount_RNA = c(NA, 500, 520, 540, 5000),
  nFeature_RNA = c(NA, 200, 210, 220, 1000),
  percent.mt = c(NA, 3, 4, 5, 30),
  row.names = paste0("cell", 1:5)
)
cell_metrics <- c(nCount_RNA = "lower", nFeature_RNA = "lower", percent.mt = "upper")
sc_report <- mad_qc_sc(cells, cell_metrics,
  transform = c(percent.mt = "none"), action = "report", verbose = FALSE)
annotated <- mad_qc_sc(cells, cell_metrics,
  transform = c(percent.mt = "none"), verbose = FALSE)
annotated$mad_qc_outlier
```

`TRUE` means any evaluated metric failed; `FALSE` means all were evaluated and
passed; `NA` means none failed but at least one was unavailable. Individual
flags stay in `sc_report$flags`. With a Seurat object, the same annotation call
adds only `mad_qc_outlier` through optional `SeuratObject` and changes no cells,
assays, reductions, or identities.

## Transformations and groups

The wrapper default is `log1p` for every selected metric. Named values are
partial overrides, so bounded rates can remain on their original scale without
repeating every count metric. Scalar `"none"` disables all transforms;
`"identity"` is its compatibility alias. Original measurements are never
modified.

```{r}
grouped <- mad_qc_bulk(bulk, metrics, transform = c(mapping_rate = "none"),
  group_by = "batch", verbose = FALSE)
attr(grouped, "mad_qc")$thresholds
```

## Inspect before filtering

```{r}
flagged[flagged$library_size_mad_outlier %in% TRUE, ]
```

This is inspection, not an automatic filtering recommendation. MAD thresholds
are adaptive heuristics rather than universal biological cutoffs. Upper RNA
count flags are not doublet calls, and doublet detection is outside package
scope. The playful console grade is presentation only, never evidence of a
biological conclusion or laboratory failure. For scientific context, see the
[scrapper manual](https://bioconductor.org/packages/release/bioc/manuals/scrapper/man/scrapper.pdf).
