---
title: "veryMAD quality-control examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{veryMAD quality-control examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(veryMAD)
```

`veryMAD` calculates explicit MAD-based QC flags on observation-level metadata.
The input measurements are preserved as `raw_value`; optional transformations
only define the calculation scale stored in `value`.

## Choosing a calculation scale

Strongly right-skewed counts, such as library size and detected-feature counts,
are often easier to threshold on a log scale. Bounded percentages, proportions,
and rates usually stay on their original scale. `veryMAD` never infers this from
column names: the transformation must be visible in the function call.

## Bulk RNA-seq sample QC

```{r bulk}
bulk_metadata <- veryMAD:::.simulate_bulk_qc_metadata(n = 240, seed = 123)

bulk_metrics <- c(
  library_size = "lower",
  detected_genes = "lower",
  mapping_rate = "lower",
  assigned_rate = "lower",
  rrna_rate = "upper"
)

bulk_transform <- c(
  library_size = "log10p",
  detected_genes = "log10p"
)

bulk_qc <- mad_qc(
  bulk_metadata,
  metrics = bulk_metrics,
  transform = bulk_transform,
  nmads = 3
)

head(bulk_qc)
summarize_mad_qc(bulk_qc, level = "metric")
```

The simulated ranges are illustrative and are not acceptance cutoffs. Duplication
rate and mitochondrial proportion are not used as universal bulk defaults here:
both can be informative in specific workflows, but their interpretation depends
on protocol and biology.

## Single-cell metadata QC

```{r single-cell}
cell_metadata <- veryMAD:::.simulate_single_cell_qc_metadata(n = 1200, seed = 123)

sc_metrics <- c(
  nCount_RNA = "both",
  nFeature_RNA = "both",
  percent.mt = "upper"
)

sc_transform <- c(
  nCount_RNA = "log10p",
  nFeature_RNA = "log10p"
)

sc_qc <- mad_qc(
  cell_metadata,
  metrics = sc_metrics,
  transform = sc_transform,
  nmads = 3
)

head(sc_qc)
summarize_mad_qc(sc_qc, level = "metric")
```

The Bioconductor single-cell workflow commonly uses lower-tail filtering for
library size and detected features. The example uses two-sided count and feature
flags so unusually high cells are visible for inspection. Those flags are not
doublet calls; doublet detection should use a dedicated method.

## Annotation

```{r annotation}
annotated_bulk <- annotate_mad_qc(bulk_metadata, bulk_qc)
head(annotated_bulk)
```

Annotation adds flag columns. It does not filter rows or change the original QC
measurements.
