---
title: "Auditable omics QC with veryMAD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditable omics QC with veryMAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(veryMAD)
```

## 1. QC data model

`mad_qc()` accepts observation-level metadata: one row is a bulk library,
single cell, spatial spot, or other observation; columns contain numeric QC
metrics and optional grouping variables. A gene-expression matrix is not QC
metadata. `group_by = NULL` uses one reference distribution per metric.

## 2. Bulk RNA-seq sample QC

The simulated ranges below are illustrative rather than acceptance cutoffs.
Condition and batch remain metadata, not automatic calculation groups.

```{r bulk}
bulk_metadata <- veryMAD:::.simulate_bulk_qc_metadata(n = 150, seed = 123)
bulk_report <- bulk_metadata |>
  mad_qc(
    metrics = c(
      library_size = "lower",
      detected_genes = "lower",
      mapping_rate = "lower",
      duplication_rate = "upper",
      percent_mitochondrial = "upper"
    ),
    transform = c(library_size = "log10p", detected_genes = "log10p")
  )
head(bulk_report)
summarize_mad_qc(bulk_report, level = "metric")
```

## 3. Single-cell QC for an individual sample

```{r cells}
cell_metadata <- veryMAD:::.simulate_single_cell_qc_metadata(n = 1000, seed = 123)
cell_report <- cell_metadata |>
  mad_qc(
    metrics = c(
      nCount_RNA = "lower",
      nFeature_RNA = "lower",
      percent.mt = "upper",
      doublet_score = "upper"
    ),
    transform = c(nCount_RNA = "log10p", nFeature_RNA = "log10p")
  )
head(cell_report)
```

All cells form one reference distribution. No sample column is invented.

## 4. Grouped QC for merged single-cell data

Grouping is optional and changes the statistical reference distribution. It can
be reasonable for independently processed samples with enough cells and strong
technical shifts, but can hide globally poor samples or important differences.

```{r merged-seurat, eval=FALSE}
# Schematic: given an existing merged Seurat object called `merged_object`.
merged_report <- mad_qc_seurat(
  merged_object,
  metrics = c(
    nCount_RNA = "lower",
    nFeature_RNA = "lower",
    percent.mt = "upper"
  ),
  transform = c(nCount_RNA = "log10p", nFeature_RNA = "log10p"),
  group_by = "orig.ident",
  action = "report"
)
```

Small groups produce unstable medians and MAD estimates. Do not group merely
because a metadata column exists.

## 5. Inspecting summaries and plots

```{r summaries}
bulk_metric_summary <- summarize_mad_qc(bulk_report, level = "metric")
bulk_sample_summary <- summarize_mad_qc(bulk_report, level = "observation")
bulk_metric_summary
head(bulk_sample_summary)
```

```{r plots, eval=FALSE}
# Requires the optional ggplot2 package.
plot_mad_qc(bulk_report, type = "distribution")
plot_mad_qc(bulk_report, type = "index")
```

Blue triangles at the panel floor represent missing, not-evaluated values.
`facet_by` changes layout only; it never changes QC thresholds.

## 6. Annotating without automatic filtering

```{r annotate}
annotated_bulk <- annotate_mad_qc(bulk_metadata, bulk_report)
head(annotated_bulk)
```

Original rows and measurements are preserved. The overall flag is `TRUE` when
any metric fails, `FALSE` only when all metrics pass, and `NA` when none fail but
at least one is not evaluated. Existing flag columns require explicit
`overwrite = TRUE`.

```{r seurat-annotation, eval=FALSE}
# Schematic: given an existing Seurat object called `sample_object`.
cell_report <- mad_qc_seurat(sample_object, action = "report")
annotated_object <- mad_qc_seurat(sample_object, action = "annotate")
```

Seurat annotation returns a modified object through R assignment semantics and
adds only intended metadata flags. It never filters cells.
