---
title: "Auditable MAD quality control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditable MAD quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(veryMAD)
```

Calculate a tidy report, inspect it, summarize the existing flags, visualize the
values and thresholds, then add reversible annotations.

```{r metadata}
metadata <- data.frame(
  sample = rep(c("A", "B"), each = 4),
  counts = c(100, 120, 130, 1000, 90, 110, 125, 900),
  percent_mt = c(2, 3, 4, 30, 2, 3, 5, 25)
)

qc <- metadata |>
  mad_qc(
    c(counts = "lower", percent_mt = "upper"),
    group_by = "sample",
    transform = c(counts = "log10p")
  )

qc
```

Summaries do not recalculate thresholds. They provide concise views of the same
auditable flags.

```{r summaries}
summarize_mad_qc(qc, level = "metric", group_by = "sample")
summarize_mad_qc(qc, level = "observation")
```

```{r plots, eval=requireNamespace("ggplot2", quietly = TRUE)}
plot_mad_qc(qc, type = "distribution", group_by = "sample")
plot_mad_qc(qc, type = "index")
```

Only after inspection, align the flags back to the original metadata.

```{r annotate}
annotate_mad_qc(metadata, qc)
```

For Seurat objects, report first and annotate only after inspection.

```{r seurat, eval=FALSE}
report <- mad_qc_seurat(object, action = "report")
report[report$is_outlier %in% TRUE, ]
object <- mad_qc_seurat(object, action = "annotate")
```

No function filters cells automatically. Keep the report with the analysis so
thresholds and exclusion decisions remain auditable.
