---
title: "Assay-agnostic MAD quality control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assay-agnostic MAD quality control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(veryMAD)
```

## 1. QC data model

`mad_qc()` accepts observation-level QC metadata: rows are observations and
columns are numeric QC metrics plus optional grouping variables. An observation
is a bulk sample or library, a single cell, or a spatial spot or cell. A full
gene-expression count matrix is not observation-level QC metadata.

With the default `group_by = NULL`, all rows form one reference distribution
for each metric. Supplying `group_by` changes the median, MAD, limits, and
outlier calls. Small groups produce unstable MAD estimates; do not group merely
because a metadata column exists.

## 2. Bulk RNA-seq sample QC

The deterministic example contains 250 libraries with realistic passing,
multi-metric failing, and missing observations. Before this step, sample-level
metrics would be calculated from the count matrix or upstream alignment reports.

```{r bulk}
bulk_metadata <- veryMAD:::.simulate_bulk_qc_metadata(n = 250, seed = 123)

bulk_qc_report <- bulk_metadata |>
  mad_qc(
    metrics = c(
      library_size = "lower",
      detected_genes = "lower",
      mapping_rate = "lower",
      duplication_rate = "upper",
      percent_mitochondrial = "upper"
    ),
    transform = c(
      library_size = "log10p",
      detected_genes = "log10p"
    )
  )

summarize_mad_qc(bulk_qc_report, level = "metric")
```

The 250 libraries—not `condition`—define one global reference distribution per
metric. Grouping by biological condition could hide condition-associated shifts
and must be an explicit analytical decision.

## 3. Single-cell QC for an individual sample

This distinct simulation contains 600 cells from one sample and no artificial
sample grouping column.

```{r single-cell}
cell_metadata <- veryMAD:::.simulate_single_cell_qc_metadata(n = 600, seed = 123)

cell_qc_report <- cell_metadata |>
  mad_qc(
    metrics = c(
      nCount_RNA = "lower",
      nFeature_RNA = "lower",
      percent.mt = "upper",
      doublet_score = "upper"
    ),
    transform = c(
      nCount_RNA = "log10p",
      nFeature_RNA = "log10p"
    )
  )

summarize_mad_qc(cell_qc_report, level = "metric")
```

All cells form one reference distribution. The equivalent Seurat report also
uses the ungrouped default:

```{r single-seurat, eval=FALSE}
cell_qc_report <- mad_qc_seurat(
  sample_object,
  metrics = c(
    nCount_RNA = "lower",
    nFeature_RNA = "lower",
    percent.mt = "upper"
  ),
  transform = c(nCount_RNA = "log10p", nFeature_RNA = "log10p"),
  action = "report"
)
```

## 4. Grouped QC for merged single-cell data

Grouping is optional and can help when independently processed samples have
strong technical differences and enough cells for stable within-sample MADs.

```{r merged-seurat, eval=FALSE}
merged_qc_report <- mad_qc_seurat(
  merged_object,
  metrics = c(
    nCount_RNA = "lower",
    nFeature_RNA = "lower",
    percent.mt = "upper"
  ),
  transform = c(nCount_RNA = "log10p", nFeature_RNA = "log10p"),
  group_by = "orig.ident",
  action = "report"
)
```

This may be appropriate when depth or mitochondrial distributions differ for
technical reasons and global thresholds unfairly penalize one sample. It may be
inappropriate when groups are small, differences are important QC signals,
grouping hides a globally poor sample, or biological groups are mistaken for
technical batches. Merged objects do not require grouping by default.

## 5. Inspecting summaries and plots

Summaries consume the saved flags; plots consume the saved values, flags, and
thresholds. Neither recalculates QC.

```{r summaries}
bulk_metric_summary <- summarize_mad_qc(bulk_qc_report, level = "metric")
bulk_sample_summary <- summarize_mad_qc(bulk_qc_report, level = "observation")

head(bulk_sample_summary)
```

```{r plots, eval=requireNamespace("ggplot2", quietly = TRUE)}
plot_mad_qc(bulk_qc_report, type = "distribution")
plot_mad_qc(bulk_qc_report, type = "index")
plot_mad_qc(bulk_qc_report, type = "index", facet_by = "condition")
```

Calculation grouping determines which rows share thresholds. `facet_by`
changes only visual arrangement. The ungrouped bulk report remains globally
thresholded even when the plot is faceted by condition.

## 6. Annotating without automatic filtering

Only after inspection, align flags back to the original metadata.

```{r annotate}
annotated_bulk_metadata <- annotate_mad_qc(bulk_metadata, bulk_qc_report)
head(annotated_bulk_metadata)
```

For Seurat objects, annotation is equally explicit and reversible:

```{r annotate-seurat, eval=FALSE}
sample_object <- mad_qc_seurat(sample_object, action = "annotate")
```

No veryMAD function automatically filters samples, libraries, cells, or spots.
